Estimating initial state and quark-gluon plasma medium properties \\ using a hybrid model with nucleon substructure \\ calibrated to p-Pb and Pb-Pb collisions at \texorpdfstring{$\mathbf{\sqrts=5.02}$}{}~TeV

### 2018-08-06 paper/substructure.tex
Final edits before posting to arXiv

\newcommand{\trento}{T\raisebox{-0.5ex}{R}ENTo}
\newcommand{\sqrts}{\sqrt{s_{NN}}}
\newcommand{\fmc}{\ensuremath{\text{fm}/c}}
\newcommand{\nch}{N_\text{ch}}
\newcommand{\ntrk}{N_\text{trk}^\text{offline}}
\newcommand{\vnk}[2]{v_#1\{#2\}}
\newcommand{\sigmaf}{\sigma_\text{fluct}}
\newcommand{\X}{\chi_\text{struct}}
\newcommand{\taufs}{\tau_\text{fs}}
\newcommand{\dmin}{d_\text{min}}
\newcommand{\Tsw}{T_\text{switch}}
\newcommand{\T}{\tilde{T}}
\newcommand{\x}{\mathbf x}
\newcommand{\y}{\mathbf y}
\newcommand{\z}{\mathbf z}
\newcommand{\paddedhline}{\noalign{\smallskip}\hline\noalign{\smallskip}}
\newcommand{\order}[1]{$\mathcal O(10^{#1})$}
\newcommand{\note}{\textcolor{theblue}{[?]}}

@@

The \trento\ model operates in the ultrarelativistic limit with a Lorentz factor $\gamma \gg 1$ such that each nucleus appears as a thin sheet of nuclear density in the laboratory frame.
The sheets of colliding nuclear density penetrate and pass through each other in proper time $\Delta \tau \approx D_\text{nucl} / (\gamma\, \beta_z)$ in the laboratory frame, where $D_\text{nucl}$ is the diameter of the nucleus in its rest frame, $\gamma$ is the usual Lorentz factor of the accelerated ions, and $\beta_z$ is their velocity along the beam axis.
The resulting nuclear overlap time $\Delta \tau \lesssim 0.1\ \fmc$ at top RHIC and LHC energies is thus sufficiently short to neglect initial transverse dynamics which occur while the nuclei pass through each other.
We therefore assume that the collision produces all secondary particles at uniform proper time $\tau = 0^+\ \fmc$, and that it deposits entropy (energy) at midrapidity which is a function of the locally varying beam integrated density of each nucleus.

@@

Within a picture of independent pairwise collisions between the constituents, a Glauber model model may be used to calculate the probability $P_\mathrm{coll}$ that the two protons collide inelastically at impact parameter $b$. In the limit when the number of constituents is large, it yields the particularly simple form
\begin{equation}
  \label{eq:pcoll}
  P_\mathrm{coll} = 1 - \exp[-\sigma_\mathrm{eff}\, T_{pp}(b)], \\
\end{equation}
where $\sigma_\mathrm{eff}$ is an effective cross section for pairwise inelastic collisions between the constituents, and $\sigma_{pp}^\mathrm{inel}$ is the total inelastic proton-proton cross section
\begin{equation}
  \label{eq:sigma_nn}
  \sigma_{pp}^\mathrm{inel} = \int 2 \pi b\, db\, P_\mathrm{coll}(b).
\end{equation}
The proton densities $\rho_{A,B}$ in Eqn.~\eqref{eq:tpp} are commonly modeled using a spherically symmetric distribution.
For instance, the original implementation of the \trento\ model uses Gaussian protons, largely because it yields a simple analytic solution to Eqn.~\eqref{eq:pcoll}.
Needless to say, such approximations are admittedly crude and may have a significant effect on the dynamics of small collision systems where the proton size is comparable to the size of the produced QGP medium.

@@

In this work, we pursue a less restrictive and more parametric description of the proton where the number of substructure degrees of freedom are uncertain as depicted in Fig.~\ref{fig:substructure}.
We model each nucleon's density $\rho_{A,B}$ as a sum of $n_c$ independent constituents
\begin{equation}
  \label{eq:rho}
  \rho_{A, B}(\x) = \frac{1}{n_c} \sum\limits_{i=1}^{n_c} \rho_c(\mathbf{x} - \mathbf{x_i}),
\end{equation}
where each constituent density $\rho_c$ is described by a Gaussian distribution of width $v$
\begin{equation}
  \label{eq:constituent_density}
  \rho_c(\mathbf{x}) = \frac{1}{(2\pi v^2)^{3/2}} \exp\left(-\frac{\x^2}{2 v^2}\right),
\end{equation}
and each constituent's position $\mathbf{x_i}$ in Eqn.~\eqref{eq:rho} is sampled from a Gaussian radial distribution with standard deviation $r$.
The effect of this additional nucleon substructure on the nuclear thickness functions is visible in Fig.~\ref{fig:thickness}

The two protons $A$, $B$ are assigned a random impact parameter, and Eqn.~\eqref{eq:pcoll} is used to sample their inelastic collision probability $P_\mathrm{coll}(b)$.
Note that this proton-proton inelastic collision probability has no direct knowledge of the individual constituent degrees of freedom; it is only \emph{indirectly} sensitive through the geometry of $\rho_{A, B}$ which depends on each of the constituent positions.
This is an important distinction between the present model and a similar nucleon substructure implementation known as the participant or ``wounded'' quark model which allows for a subset of quarks (constituents) to participate inside a single nucleon \cite{ANISOVICH1978477, Broniowski:2016pvx}.
The proton, unlike the nucleus, cannot produce semi-stable spectator fragments in a high-energy collision.
Any spectator quarks produced by a wounded quark model would be colored objects that necessarily contribute to secondary particle production as they fragment and recombine to form color-neutral hadrons.
We correspondingly require that the nucleons in Eqn.~\eqref{eq:rho} participate as singular objects, such that all spectator matter discarded by the simulation is appropriately color-neutral and inert.

Assuming our two protons collide at the sampled impact parameter $\mathbf{b}$, we assign each a \emph{fluctuated} thickness
\begin{equation}
  \label{eq:fluctuated_thick}
  \T_{A, B}(\x) \equiv \int dz\, \frac{1}{n_c} \sum\limits_{i=1}^{n_c} \gamma_i\, \rho_c \,(\mathbf{x} - \mathbf{x_i} \pm \mathbf{b}/2),
\end{equation}
equal to the beam-integrated proton density in Eqn.~\eqref{eq:rho}, with each constituent shifted by the appropriate impact parameter offset, and multiplied by a gamma random variable $\gamma_i$ with unit mean and variance $1/k$.
These \emph{ad hoc} gamma random weights are necessary to describe the large fluctuations observed in minimum bias proton-proton collisions, although their exact physical origin is not well understood.

Within the eikonal approximation, the initial entropy deposited at midrapidity at proper time $\tau=0^+\ \fmc$ can be represented by some mapping ${f: \T_A, \T_B \mapsto dS/dy}$.
A natural first guess for the mapping is the arithmetic mean
\begin{equation}
  \frac{dS}{dy}\bigg\vert_{y=0} \propto \frac{\T_A + \T_B}{2}
\end{equation}
of the participant nuclear matter $\T_A$ and $T_B$, which is simply a wounded nucleon model up to meaningless factor of two in the normalization.
The wounded nucleon model was in fact one of the first such mappings used as a proxy for initial particle production and entropy deposition in relativistic heavy-ion collisions \cite{Bialas:1976ed}.
It was quickly realized, however, that the wounded nucleon model predicts the wrong scaling for charged particle production as a function of collision centrality and hence the wrong scaling for initial entropy deposition as a function of participant thicknesses $\T_A$ and $\T_B$ \cite{Kharzeev:2000ph}.

A simple remedy is to replace the arithmetic mean of the wounded nucleon model with a more flexible parametrization
\begin{equation}
  \label{eq:gmean}
  \frac{dS}{dy}\bigg\vert_{y=0} \propto \left( \frac{\T_A^p + \T_B^p}{2} \right)^{1/p},
\end{equation}
based on a family of functions known as the generalized mean(s).
This parametrization introduces a dimensionless parameter $p$ which varies the scaling behavior of initial entropy deposition at midrapidity.
For certain discrete values of $p$, it reduces to well known functional forms such as the arithmetic, geometric and harmonic means:
\begin{equation}
  \newlength{\extraspace}
  \setlength{\extraspace}{0.5ex}
  \frac{dS}{dy}\bigg\vert_{y=0} \propto
  \begin{cases}
    \max(T_A, T_B) & p \rightarrow +\infty, \\[\extraspace]
    (T_A + T_B)/2 & p = +1, \hfill \text{ (arithmetic)} \\[\extraspace]
    \sqrt{T_A T_B} & p = 0, \hfill \text{ (geometric)} \\[\extraspace]
    2\, T_A T_B/(T_A + T_B) & p = -1, \hfill \text{ (harmonic)} \\[\extraspace]
    \min(T_A, T_B) & p \rightarrow -\infty.
  \end{cases}
  \label{eq:trento_p}
\end{equation}
Conveniently, it has been shown that the generalized mean ansatz is able to mimic the scaling behavior of certain saturation based initial condition models \cite{Bernhard:2016tnd}, and hence it should serve as a reasonable parametric form for exploring QGP entropy deposition, assuming imperfect knowledge of saturation effects in nature.
Of course the model is \emph{not} a substitute for first principle theory calculations, and it may fail to reproduce nuanced features of \emph{ab initio} models such as the existence of short-range gluon field fluctuations \cite{Schenke:2012wb}.

@@

Equation~\eqref{eq:gmean} is a purely local function of nuclear density in the transverse plane and should (in principle) be equally valid for any pair of colliding nuclei at sufficiently high beam energy.
The model readily generalizes from individual proton-proton collisions to arbitrary nucleus-nucleus collisions by summing the participant thicknesses $\T_{A,B}$ over all nucleons which participate in one or more inelastic collisions.
The only modeling difference between $p$-$p$ collisions and Pb-Pb collisions is the number and the position of the nucleons.

@@

The parametric entropy deposition ansatz in Eqn.~\eqref{eq:gmean} does not provide any information about the initial masses or momenta of particles produced in the collision.
In general, these details will affect the dynamics predicted by the collisionless Boltzmann equation
\begin{equation}
  \label{freestream}
  p^\mu \partial_\mu f(x, p) = 0,
\end{equation}
through its dependence on the underlying distribution function $f(x, p)$, and hence are necessary inputs for any free-streaming implementation.
Equation~\eqref{freestream}, however, simplifies for massless partons with momentum distributions that are locally isotropic.
Under this assumption, it was shown in Refs.~\cite{Broniowski:2008qk, Liu:2015nwa} that the energy-momentum tensor $T^{\mu\nu}$ of partons at midrapidity and time $\tau$ only depends on its spatial distribution at some earlier time $\tau_0$, but not its $p_\perp$-distribution, which could vary as a function of position.

We thus follow the procedure of the authors and reinterpret our parametrization for the initial transverse entropy density at midrapidity $s(\mathbf{x_\perp}, \tau_0)\vert_{\eta=0}$ as a parametrization for the initial transverse density of massless noninteracting partons $n(\mathbf{x}, \tau_0)\vert_{\eta=0}$.
The free-streamed energy-momentum tensor $T^{\mu\nu}(x, y, \eta, \tau)$ at transverse coordinate $(x, y)$, midrapidity $\eta=0$, and proper time $\tau > \tau_0$ is then given by
\begin{multline}
  \label{energy-momentum}
  T^{\mu\nu}(x, y, 0, \tau) = \frac{1}{\tau} \int_{0}^{2\pi} d\phi\, n(x - \Delta \tau \cos \phi, y - \Delta \tau \sin \phi) \\ \times
  \begin{bmatrix}
    1 & \cos\phi & \sin\phi  & 0\\
    \cos\phi & \cos^2\phi & \cos\phi\sin\phi & 0 \\
    \sin\phi & \sin\phi \cos\phi & \sin^2\phi & 0\\
    0 & 0 & 0 & 0
  \end{bmatrix},
\end{multline}
where $\Delta\tau$ is the free-streaming time $\Delta\tau = \tau - \tau_0$.
The solution \eqref{energy-momentum} is decomposed in hydrodynamic form
\begin{equation}
  \label{hydro-eqn}
  T^{\mu\nu} = e u^\mu u^\nu - (P + \Pi) \Delta^{\mu\nu} + \pi^{\mu\nu},
\end{equation}
where $e$ and $P$ are the energy density and pressure in the local fluid rest frame, $u^\mu$ is the local fluid velocity, ${\Delta^{\mu\nu} \equiv g^{\mu\nu} - u^\mu u^\nu}$ is the projector onto the space orthogonal to $u^\mu$, and $\Pi$ and $\pi^{\mu\nu}$ are the bulk pressure and shear stress tensor respectively.
We then solve for the energy density $e$ and fluid velocity $u^\mu$ using the Landau matching condition which defines the fluid rest frame velocity as the time-like eigenvector of $T^{\mu\nu}$ with energy density $e$ as its eigenvalue,
\begin{equation}
  T^{\mu\nu} u_\nu = e u^\mu.
\end{equation}
The initial bulk and shear corrections are finally solved for by subtracting the ideal pressure from the total pressure to find $\Pi$, then solving for $\pi^{\mu\nu}$ using Eqn.~\eqref{hydro-eqn}
\begin{align}
  \Pi &= -\frac{1}{3} \mathrm{Tr}(\Delta_{\mu\nu} T^{\mu\nu}) - P,\\
  \pi^{\mu\nu} &= T^{\mu\nu} - e u^\mu u^\nu + (P + \Pi) \Delta^{\mu\nu}.
\end{align}

@@

After free-streaming for proper time $\taufs$, we transition to viscous hydrodynamics which solves the conservation equations
\begin{equation}
  \label{eq:continuity}
  \partial_\mu T^{\mu\nu} = 0
\end{equation}
for the hydrodynamic energy-momentum tensor $T^{\mu\nu}$ expressed in Eqn.~\eqref{hydro-eqn} using a set of second-order Israel-Stewart equations formulated in the 14-moment approximation
\cite{Israel:1979wp, Israel:1976aa, Denicol:2012cn, Denicol:2010xn}.
This produces a pair of relaxation-type equations
\begin{subequations}
  \label{eq:relaxation}
  \begin{align}
    \tau_\Pi \Pi + \dot{\Pi} &=
      - \zeta \theta - \delta_{\Pi\Pi} \Pi\theta
      + \lambda_{\Pi\pi} \pi^{\mu\nu} \sigma_{\mu\nu}, \\[1ex]
    \tau_\pi \dot{\pi}^{\langle \mu\nu \rangle} + \pi^{\mu\nu} &=
      2\eta\sigma^{\mu\nu} - \delta_{\pi\pi} \pi^{\mu\nu} \theta
      + \phi_7 \pi_\alpha^{\langle \mu} \pi^{\nu \rangle \alpha} \nonumber \\
      &\qquad {} - \tau_{\pi\pi} \pi_\alpha^{\langle \mu}\sigma^{\nu \rangle \alpha}
      + \lambda_{\pi\Pi} \Pi \sigma^{\mu\nu},
  \end{align}
\end{subequations}
for the bulk pressure $\Pi$ and shear-stress $\pi^{\mu\nu}$.
We model the shear viscosity $\eta$ and bulk viscosity $\zeta$ as unknown temperature dependent quantities and fix the remaining transport coefficients $\{\tau_\Pi, \delta_{\Pi\Pi}, \lambda_{\Pi\pi}, \tau_\pi, \delta_{\pi\pi}, \phi_7, \tau_{\pi\pi}, \lambda_{\pi\Pi}\}$ using analytic results derived in the limit of small but finite masses \cite{Denicol:2014vaa}.

@@

We parametrize the temperature dependence of the QGP viscosities in order to marginalize over their uncertainty when calibrating to data.
For the specific shear viscosity $\eta/s$, we use a modified linear ansatz
\begin{equation}
  \label{eq:shear_viscosity}
  (\eta/s)(T) = (\eta/s)_\mathrm{min} + (\eta/s)_\mathrm{slope}\cdot(T - T_c)\cdot(T/T_c)^{(\eta/s)_\mathrm{crv}},
\end{equation}
where $\eta/s$ min, slope and curvature are tunable parameters and $T_c=0.154$~GeV is the pseudocritical transition temperature of the HotQCD EoS.
While for the specific bulk viscosity $\zeta/s$, we use an unnormalized Cauchy distribution
\begin{equation}
  \label{eq:bulk_viscosity}
  (\zeta/s)(T) = \frac{(\zeta/s)_\mathrm{max}}{1 + \left(\dfrac{T - (\zeta/s)_{T_0}}{(\zeta/s)_\mathrm{width}}\right)^2},
\end{equation}
described by tunable maximum, width, and location ($T_0$) parameters.
Figure~\ref{fig:viscosity_dof} shows several of the possible curves parametrized by Eqns.~\eqref{eq:shear_viscosity} and \eqref{eq:bulk_viscosity}, although many more are possible.

@@

We evolve the system hydrodynamically down to a pre-specified switching isotherm $T_{sw}$ at which point the medium is converted into particles using the Cooper-Frye formula \cite{PhysRevD.10.186}
\begin{equation}
  \label{cooper-frye}
  E \frac{dN_i}{d^3p} = \frac{g_i}{(2\pi)^3} \int_\Sigma f_i(x, p)\, p^\mu\, d^3\sigma_\mu,
\end{equation}
where $i$ is an index over species, $f_i$ is the distribution function of that species, and $d^3\sigma_\mu$ is a volume element of the isothermal hypersurface $\Sigma$ defined by $T_\mathrm{sw}$.
Thermal particles are then sampled in the rest frame of each fluid cell according to a Bose-Einstein or Fermi-Dirac distribution at zero baryon chemical potential
\begin{equation}
  \label{distribution}
  f(m, p) = \frac{1}{\exp(\sqrt{m^2 + p^2}/T) \mp 1},
\end{equation}
where $m$ is the mass of the sampled particle, $p$ is its momentum, and $T$ is the temperature of the fluid cell.

>>>>

\newcommand{\trento}{T\raisebox{-0.5ex}{R}ENTo}
\newcommand{\sqrts}{\sqrt{s_\mathrm{NN}}}
\newcommand{\fmc}{\ensuremath{\text{fm}/c}}
\newcommand{\nch}{N_\text{ch}}
\newcommand{\ntrk}{N_\text{trk}^\text{offline}}
\newcommand{\vnk}[2]{v_#1\{#2\}}
\newcommand{\sigmaf}{\sigma_\text{fluct}}
\newcommand{\X}{\chi_\text{struct}}
\newcommand{\taufs}{\tau_\text{fs}}
\newcommand{\dmin}{d_\text{min}}
\newcommand{\Tsw}{T_\text{switch}}
\newcommand{\T}{\tilde{T}}
\newcommand{\x}{\mathbf x}
\newcommand{\y}{\mathbf y}
\newcommand{\z}{\mathbf z}
\newcommand{\paddedhline}{\noalign{\smallskip}\hline\noalign{\smallskip}}
\newcommand{\order}[1]{$\mathcal O(10^{#1})$}
\newcommand{\note}{\textcolor{theblue}{[?]}}

@@

The \trento\ model operates in the ultrarelativistic limit with a Lorentz factor $\gamma \gg 1$ such that each nucleus appears as a thin sheet of nuclear density in the laboratory frame.
The sheets of colliding nuclear density penetrate and pass through each other in proper time $\Delta \tau \approx D_\text{nucl} / (\gamma\, \beta_z)$ in the laboratory frame, where $D_\text{nucl}$ is the diameter of the nucleus in its rest frame, $\gamma$ is the usual Lorentz factor of the accelerated ions, and $\beta_z$ is their velocity along the beam axis.
The resulting nuclear overlap time $\Delta \tau \lesssim 0.1\ \fmc$ at top RHIC and LHC energies, and thus it is safe to neglect the initial transverse dynamics which occur while the nuclei pass through each other.
We therefore assume that the collision produces all secondary particles at uniform proper time $\tau = 0^+\ \fmc$, and that it deposits entropy (energy) at midrapidity which is a function of the locally varying beam integrated density of each nucleus.

@@

Within a picture of independent pairwise collisions between the constituents, a Glauber model model may be used to calculate the probability $P_\mathrm{coll}$ that the two protons collide inelastically at impact parameter $b$. In the limit when the number of constituents is large, it yields the particularly simple form
\begin{equation}
  \label{eq:pcoll}
  P_\mathrm{coll} = 1 - \exp[-\sigma_\mathrm{eff}\, T_{pp}(b)], \\
\end{equation}
where $\sigma_\mathrm{eff}$ is an effective cross section for pairwise inelastic collisions between the constituents, and $\sigma_{pp}^\mathrm{inel}$ is the total inelastic proton-proton cross section
\begin{equation}
  \label{eq:sigma_nn}
  \sigma_{pp}^\mathrm{inel} = \int 2 \pi b\, db\, P_\mathrm{coll}(b).
\end{equation}
The proton densities $\rho_{A,B}$ in Eq.~\eqref{eq:tpp} are commonly modeled using a spherically symmetric distribution.
For instance, the original implementation of the \trento\ model uses Gaussian protons, largely because it yields a simple analytic solution to Eq.~\eqref{eq:pcoll}.
Needless to say, such approximations are admittedly crude and may have a significant effect on the dynamics of small collision systems where the proton size is comparable to the size of the produced QGP medium.

@@

In this work, we pursue a less restrictive and more parametric description of the proton where the number of substructure degrees of freedom are uncertain as depicted in Fig.~\ref{fig:substructure}.
We model each nucleon's density $\rho_{A,B}$ as a sum of $n_c$ independent constituents
\begin{equation}
  \label{eq:rho}
  \rho_{A, B}(\x) = \frac{1}{n_c} \sum\limits_{i=1}^{n_c} \rho_c(\mathbf{x} - \mathbf{x_i}),
\end{equation}
where each constituent density $\rho_c$ is described by a Gaussian distribution of width $v$
\begin{equation}
  \label{eq:constituent_density}
  \rho_c(\mathbf{x}) = \frac{1}{(2\pi v^2)^{3/2}} \exp\left(-\frac{\x^2}{2 v^2}\right),
\end{equation}
and each constituent's position $\mathbf{x_i}$ in Eq.~\eqref{eq:rho} is sampled from a Gaussian radial distribution with standard deviation $r$.
The effect of this additional nucleon substructure on the nuclear thickness functions is visible in Fig.~\ref{fig:thickness}

@@

The two protons $A$, $B$ are assigned a random impact parameter, and Eq.~\eqref{eq:pcoll} is used to sample their inelastic collision probability $P_\mathrm{coll}(b)$.
Note that this proton-proton inelastic collision probability has no direct knowledge of the individual constituent degrees of freedom; it is only \emph{indirectly} sensitive through the geometry of $\rho_{A, B}$ which depends on each of the constituent positions.
This is an important distinction between the present model and a similar nucleon substructure implementation known as the participant or ``wounded'' quark model which allows for a subset of quarks (constituents) to participate inside a single nucleon \cite{ANISOVICH1978477, Broniowski:2016pvx}.
The proton, unlike the nucleus, cannot produce semi-stable spectator fragments in a high-energy collision.
Any spectator quarks produced by a wounded quark model would be colored objects that necessarily contribute to secondary particle production as they fragment and recombine to form color-neutral hadrons.
We correspondingly require that the nucleons in Eq.~\eqref{eq:rho} participate as singular objects, such that all spectator matter discarded by the simulation is appropriately color-neutral and inert.

Assuming our two protons collide at the sampled impact parameter $\mathbf{b}$, we assign each a \emph{fluctuated} thickness
\begin{equation}
  \label{eq:fluctuated_thick}
  \T_{A, B}(\x) \equiv \int dz\, \frac{1}{n_c} \sum\limits_{i=1}^{n_c} \gamma_i\, \rho_c \,(\mathbf{x} - \mathbf{x_i} \pm \mathbf{b}/2),
\end{equation}
equal to the beam-integrated proton density in Eq.~\eqref{eq:rho}, with each constituent shifted by the appropriate impact parameter offset, and multiplied by a gamma random variable $\gamma_i$ with unit mean and variance $1/k$.
These \emph{ad hoc} gamma random weights are necessary to describe the large fluctuations observed in minimum bias proton-proton collisions, although their exact physical origin is not well understood.

Within the eikonal approximation, the initial entropy deposited at midrapidity and proper time $\tau=0^+\ \fmc$ is some function
\begin{equation}
  f: \T_A, \T_B \mapsto \frac{dS}{d^2x_\perp d\eta} \bigg\vert_{\eta=0},
\end{equation}
of the local density of participant matter $\T_A$, $\T_B$ in each nucleus.
A natural first guess for this mapping is the arithmetic mean
\begin{equation}
  \frac{dS}{d^2x_\perp d\eta}\,\bigg\vert_{\eta=0} \propto \frac{\T_A + \T_B}{2},
\end{equation}
which is simply a wounded nucleon model up to meaningless factor of two in the normalization.
The wounded nucleon model was in fact one of the first such mappings used as a proxy for initial particle production and entropy deposition in relativistic heavy-ion collisions \cite{Bialas:1976ed}.
It was quickly realized, however, that the wounded nucleon model predicts the wrong scaling for charged particle production as a function of collision centrality and hence the wrong scaling for initial entropy deposition as a function of participant thicknesses $\T_A$ and $\T_B$ \cite{Kharzeev:2000ph}.

A simple remedy is to replace the arithmetic mean of the wounded nucleon model with a more flexible parametrization
\begin{equation}
  \label{eq:gmean}
  \frac{dS}{d^2x_\perp d\eta}\,\bigg\vert_{\eta=0} \propto \bigg( \frac{\T_A^p + \T_B^p}{2} \bigg)^{1/p},
\end{equation}
based on a family of functions known as the generalized mean(s).
This parametrization introduces a dimensionless parameter $p$ which varies the scaling behavior of initial entropy deposition at midrapidity.
For certain discrete values of $p$, it reduces to well known functional forms such as the arithmetic, geometric and harmonic means:
\begin{equation}
  \newlength{\extraspace}
  \setlength{\extraspace}{0.5ex}
  \frac{dS}{d\eta}\bigg\vert_{\eta=0} \propto
  \begin{cases}
    \max(T_A, T_B) & p \rightarrow +\infty, \\[\extraspace]
    (T_A + T_B)/2 & p = +1, \hfill \text{ (arithmetic)} \\[\extraspace]
    \sqrt{T_A T_B} & p = 0, \hfill \text{ (geometric)} \\[\extraspace]
    2\, T_A T_B/(T_A + T_B) & p = -1, \hfill \text{ (harmonic)} \\[\extraspace]
    \min(T_A, T_B) & p \rightarrow -\infty.
  \end{cases}
  \label{eq:trento_p}
\end{equation}
Conveniently, it has been shown that the generalized mean ansatz is able to mimic the scaling behavior of certain saturation based initial condition models \cite{Bernhard:2016tnd}, and hence it should serve as a reasonable parametric form for exploring QGP entropy deposition, assuming imperfect knowledge of saturation effects in nature.
Of course the model is \emph{not} a substitute for first principle theory calculations, and it may fail to reproduce nuanced features of \emph{ab initio} models such as the existence of short-range gluon field fluctuations \cite{Schenke:2012wb}.

Equation~\eqref{eq:gmean} is a purely local function of nuclear density in the transverse plane and should (in principle) be equally valid for any pair of colliding nuclei at sufficiently high beam energy.
The model readily generalizes from individual proton-proton collisions to arbitrary nucleus-nucleus collisions by summing the participant thicknesses $\T_{A,B}$ over all nucleons which participate in one or more inelastic collisions.
The only modeling difference between $p$-$p$, $p$-Pb, and Pb-Pb collisions is the number and the position of the nucleons.

@@

The parametric entropy deposition ansatz in Eq.~\eqref{eq:gmean} does not provide any information about the initial masses or momenta of particles produced in the collision.
In general, these details will affect the dynamics predicted by the collisionless Boltzmann equation
\begin{equation}
  \label{freestream}
  p^\mu \partial_\mu f(x, p) = 0,
\end{equation}
through its dependence on the underlying distribution function $f(x, p)$, and hence are necessary inputs for any free-streaming implementation.
Equation~\eqref{freestream}, however, simplifies for massless partons with momentum distributions that are locally isotropic.
Under this assumption, it was shown in Refs.~\cite{Broniowski:2008qk, Liu:2015nwa} that the energy-momentum tensor $T^{\mu\nu}$ of partons at midrapidity and time $\tau$ only depends on its spatial distribution at some earlier time $\tau_0$, but not its $p_\perp$-distribution, which could vary as a function of position.

The entropy density for a gas of massless noninteracting particles is very nearly proportional to its particle density, and so we are able to recast the parametrization in Eq.~\eqref{eq:gmean} in the form
\begin{equation}
  \frac{dN}{d^2x_\perp d\eta}\,\bigg\vert_{\eta=0} = \, \text{Norm} \times \bigg(\frac{\T_A^p + \T_B^p}{2}\bigg)^{1/p},
\end{equation}
where the left-side is the initial density of free-streaming partons at midrapidity.
The resulting free-streamed energy-momentum tensor $T^{\mu\nu}(x, y, \eta, \tau)$ at transverse coordinate $(x, y)$, midrapidity $\eta=0$, and proper time $\tau > \tau_0$ is then given by
\begin{multline}
  \label{energy-momentum}
  T^{\mu\nu}(x, y, 0, \tau) = \frac{1}{\tau} \int_{0}^{2\pi} d\phi\, n(x - \Delta \tau \cos \phi, y - \Delta \tau \sin \phi) \\ \times
  \begin{bmatrix}
    1 & \cos\phi & \sin\phi  & 0\\
    \cos\phi & \cos^2\phi & \cos\phi\sin\phi & 0 \\
    \sin\phi & \sin\phi \cos\phi & \sin^2\phi & 0\\
    0 & 0 & 0 & 0
  \end{bmatrix},
\end{multline}
where $n$ is the local density of massless partons, and $\Delta\tau$ is the free-streaming time $\Delta\tau = \tau - \tau_0$.
The solution \eqref{energy-momentum} can then be decomposed in hydrodynamic form
\begin{equation}
  \label{hydro-eqn}
  T^{\mu\nu} = e u^\mu u^\nu - (P + \Pi) \Delta^{\mu\nu} + \pi^{\mu\nu},
\end{equation}
where $e$ and $P$ are the energy density and pressure in the local fluid rest frame, $u^\mu$ is the local fluid velocity, ${\Delta^{\mu\nu} \equiv g^{\mu\nu} - u^\mu u^\nu}$ is the projector onto the space orthogonal to $u^\mu$, and $\Pi$ and $\pi^{\mu\nu}$ are the bulk pressure and shear stress tensor respectively.
We then solve for the energy density $e$ and fluid velocity $u^\mu$ using the Landau matching condition which defines the fluid rest frame velocity as the time-like eigenvector of $T^{\mu\nu}$ with energy density $e$ as its eigenvalue,
\begin{equation}
  T^{\mu\nu} u_\nu = e u^\mu.
\end{equation}
The initial bulk and shear corrections are finally solved for by subtracting the ideal pressure from the total pressure to find $\Pi$, then solving for $\pi^{\mu\nu}$ using Eq.~\eqref{hydro-eqn}
\begin{align}
  \Pi &= -\frac{1}{3} \mathrm{Tr}(\Delta_{\mu\nu} T^{\mu\nu}) - P,\\
  \pi^{\mu\nu} &= T^{\mu\nu} - e u^\mu u^\nu + (P + \Pi) \Delta^{\mu\nu}.
\end{align}

@@

After free-streaming for proper time $\taufs$, we transition to viscous hydrodynamics which solves the conservation equations
\begin{equation}
  \label{eq:continuity}
  \partial_\mu T^{\mu\nu} = 0
\end{equation}
for the hydrodynamic energy-momentum tensor $T^{\mu\nu}$ expressed in Eq.~\eqref{hydro-eqn} using a set of second-order Israel-Stewart equations formulated in the 14-moment approximation
\cite{Israel:1979wp, Israel:1976aa, Denicol:2012cn, Denicol:2010xn}.
This produces a pair of relaxation-type equations
\begin{subequations}
  \label{eq:relaxation}
  \begin{align}
    \tau_\Pi \Pi + \dot{\Pi} &=
      - \zeta \theta - \delta_{\Pi\Pi} \Pi\theta
      + \lambda_{\Pi\pi} \pi^{\mu\nu} \sigma_{\mu\nu}, \\[1ex]
    \tau_\pi \dot{\pi}^{\langle \mu\nu \rangle} + \pi^{\mu\nu} &=
      2\eta\sigma^{\mu\nu} - \delta_{\pi\pi} \pi^{\mu\nu} \theta
      + \phi_7 \pi_\alpha^{\langle \mu} \pi^{\nu \rangle \alpha} \nonumber \\
      &\qquad {} - \tau_{\pi\pi} \pi_\alpha^{\langle \mu}\sigma^{\nu \rangle \alpha}
      + \lambda_{\pi\Pi} \Pi \sigma^{\mu\nu},
  \end{align}
\end{subequations}
for the bulk pressure $\Pi$ and shear-stress $\pi^{\mu\nu}$.
We model the shear viscosity $\eta$ and bulk viscosity $\zeta$ as unknown temperature dependent quantities and fix the remaining transport coefficients $\{\tau_\Pi, \delta_{\Pi\Pi}, \lambda_{\Pi\pi}, \tau_\pi, \delta_{\pi\pi}, \phi_7, \tau_{\pi\pi}, \lambda_{\pi\Pi}\}$ using analytic results derived in the limit of small but finite masses \cite{Denicol:2014vaa}.

@@

We parametrize the temperature dependence of the QGP viscosities in order to marginalize over their uncertainty when calibrating to data.
For the specific shear viscosity $\eta/s$, we use a modified linear ansatz
\begin{equation}
  \label{eq:shear_viscosity}
  (\eta/s)(T) = (\eta/s)_\mathrm{min} + (\eta/s)_\mathrm{slope}\cdot(T - T_c)\cdot(T/T_c)^{(\eta/s)_\mathrm{crv}},
\end{equation}
where $\eta/s$ min, slope and curvature are tunable parameters and $T_c=0.154$~GeV is the pseudocritical transition temperature of the HotQCD EoS.
While for the specific bulk viscosity $\zeta/s$, we use an unnormalized Cauchy distribution
\begin{equation}
  \label{eq:bulk_viscosity}
  (\zeta/s)(T) = \frac{(\zeta/s)_\mathrm{max}}{1 + \left(\dfrac{T - (\zeta/s)_{T_0}}{(\zeta/s)_\mathrm{width}}\right)^2},
\end{equation}
described by tunable maximum, width, and location ($T_0$) parameters.
Figure~\ref{fig:viscosity_dof} shows several of the possible curves parametrized by Eqs.~\eqref{eq:shear_viscosity} and \eqref{eq:bulk_viscosity}, although many more are possible.

@@

We evolve the system hydrodynamically down to a pre-specified switching isotherm $T_{sw}$ at which point the medium is converted into particles using the Cooper-Frye formula \cite{PhysRevD.10.186}
\begin{equation}
  \label{cooper-frye}
  E \frac{dN_i}{d^3p} = \frac{g_i}{(2\pi)^3} \int_\Sigma f_i(x, p)\, p^\mu\, d^3\sigma_\mu,
\end{equation}
where $i$ is an index over species, $f_i$ is the distribution function of that species, and $d^3\sigma_\mu$ is a volume element of the isothermal hypersurface $\Sigma$ defined by $\Tsw$.
Thermal particles are then sampled in the rest frame of each fluid cell according to a Bose-Einstein or Fermi-Dirac distribution at zero baryon chemical potential
\begin{equation}
  \label{distribution}
  f(m, p) = \frac{1}{\exp(\sqrt{m^2 + p^2}/T) \mp 1},
\end{equation}
where $m$ is the mass of the sampled particle, $p$ is its momentum, and $T$ is the temperature of the fluid cell.
